---
title: "SIM - Assignment 1"
author: "Míriam Méndez, Gabriel Zarate"
date: \today
output: pdf_document
subtitle: "Medical cost"
editor_options: 
  chunk_output_type: console
---

*Loading Libraries*

```{r}
# Load Required Packages: to be increased over the course

requiredPackages <- c("missMDA","effects","FactoMineR","car","factoextra","RColorBrewer","ggplot2","dplyr","ggmap","ggthemes","knitr","chemometrics","rpart","ROCR","corrr","readxl","RColorBrewer","psych","corrplot","plotly","xlsx","reshape2","scales","stargazer","kableExtra","lmtest")

package.check <- lapply(requiredPackages, FUN = function(x) {
  if (!require(x, character.only = TRUE)) {
    install.packages(x, dependencies = TRUE)
    library(x, character.only = TRUE)
  }
})

```

*Reading Data*

```{r}
df <- read.csv("insurance.csv")
summary(df)

View(df)
```

# Data Preparation

## Remove duplicates

```{r}
df[duplicated(df), ]
nrow(df)

df <- df[!duplicated(df), ]
nrow(df)
```

## Fixing Structural Errors

```{r}

```

## Check Data Types

```{r}
lapply(df, class)
categ_cols <- c( 'sex', 'smoker', 'region')
df[categ_cols] = lapply(df[categ_cols], FUN = as.factor)
lapply(df, class)

```

Univariate Outliers

```{r}
par(mfrow=c(1,2))

Boxplot(df[,c(1,3,4)])
#Age no outliers
#Children no outliers

Boxplot(df$bmi)
Boxplot(df$charges)
#bmi and charges seem to have outliers

#Checking outliers for bmi

ss<-summary(df$bmi);ss
# Upper/lower severe threshold
utso2<-ss[5]+3*(ss[5]-ss[2]);utso2
utsi2<-ss[2]-3*(ss[5]-ss[2]);utsi2
# Upper/lower mild threshold
utmo2<-ss[5]+1.5*(ss[5]-ss[2]);utmo2
utmi2<-ss[2]-1.5*(ss[5]-ss[2]);utmi2

Boxplot(df$bmi,id=list(n=Inf,labels=row.names(df)))
Boxplot(df$bmi)
abline(h=utso2,col="red",lwd=3)
abline(h=utsi2,col="red",lwd=3)
abline(h=utmo2,col="green",lwd=3)
abline(h=utmi2,col="green",lwd=3)

lls.bmi<-which((df$bmi>utso2)|(df$bmi<utsi2));lls
df[lls.bmi,]
llm.bmi<-which((df$bmi>utmo2)|(df$bmi<utmi2));llm
df[llm.bmi,]

par(mfrow=c(1,1))

#Checking outliers for charges

ss<-summary(df$charges);ss
# Upper/lower severe threshold
utso2<-ss[5]+3*(ss[5]-ss[2]);utso2
utsi2<-ss[2]-3*(ss[5]-ss[2]);utsi2
# Upper/lower mild threshold
utmo2<-ss[5]+1.5*(ss[5]-ss[2]);utmo2
utmi2<-ss[2]-1.5*(ss[5]-ss[2]);utmi2

Boxplot(df$charges,id=list(n=Inf,labels=row.names(df)))
Boxplot(df$charges)
abline(h=utso2,col="red",lwd=3)
abline(h=utsi2,col="red",lwd=3)
abline(h=utmo2,col="green",lwd=3)
abline(h=utmi2,col="green",lwd=3)

lls<-which((df$charges>utso2)|(df$charges<utsi2));lls
df[lls,]
llm<-which((df$charges>utmo2)|(df$charges<utmi2));llm
df[llm,]

par(mfrow=c(1,1))

#Setting severe outliers from charges as NA

df[lls,"charges"]<-NA

summary(df$charges)
```

Check missing data

```{r}
mis_col = colSums(is.na(df))
mis_col

#Only charges has missing data (outliers)

#They can't be imputed because it is the target variable, so they are deleted

md<-which(is.na(df$charges));md
df <- df[-md,]
nrow(df)

```

Multivariate Outliers

```{r}
library(chemometrics)
res.mout <- Moutlier( df[ , c(1,3,4,7)], quantile = 0.999 )

par(mfrow=c(1,1))
plot( res.mout$md, res.mout$rd )
abline( h=res.mout$cutoff, lwd=2, col="red")
abline( v=res.mout$cutoff, lwd=2, col="red")

llmout <- which( ( res.mout$md > res.mout$cutoff ) & (res.mout$rd > res.mout$cutoff) );llmout
df[llmout,]
res.mout$md[llmout]

#Since there is only one multivariate outlier, we delete it
df <- df[-llmout,]

```

Data Validation

```{r}
summary(df)

#Checking bmi mild outliers
df[llm.bmi,]

#There are two cases we have decided to delete because they aren't common, they are on the obbessity type IV, and have a really low age

df <- df[-c(844,1309),]

```

# Second part

## Create factors for qualitative variables [Done]

```{r}
summary(df)

str(df)
```

## Determine if the response variable (charges) has an acceptably normal distribution

The distribution of charges is right-skewed. We can confirm this visually using a histogram:

```{r}

#Normal Check
hist(df$charges, freq=F, main="charges")
curve(dnorm(x,mean(x),sd(x)),lwd=2,add=T,col="red")
shapiro.test(df$charges)

#Log normal check
hist(log(df$charges), freq=F, main="charges")
curve(dnorm(x,mean(x),sd(x)),lwd=2,add=T,col="red")
shapiro.test(log(df$charges))

```

It does not have an acceptable normal distribution. Most of the individuals spend between \$0 and \$15,000 on health insurance, although the tail of the distribution extends far these peaks.

## Address test to discard serial correlation [falta]

You have seen acf() method to plot autocorrelation in calories. The plot clearly shows a
serial correlation. Durbin-Watson test is an inferential tool that addresses whether
autocorrelation is present and the null hypothesis is clearly rejected, thus again,
autocorrelation is present in the cal column of the pizza dataset.

```{r}

# autocorrelation
acf(df$charges)

# Durbin-Watson Test for serial correlation
dwtest(df$charge ~ 1)

```

The autocorrelation is significantly different from zero at 5% level and we can see that there are periods of low and high variance.

## Detect univariant and multivariant outliers [Done]

## Preliminary exploratory analysis

```{r}

pairs.panels(df[c(1,3,4,7)])
```

Non of the correlations in the matrix are considered strong, whereas all have a positive correlation, meaning that as *feature 1* increase, *feature 2* also increase.

In these scatter plots we have found interesting relationships between:

-   *age* and *charges* : displays several relatively straight lines.

-   *bmi* and *charges*: has two distinct groups of points.

-   
